dispRity manual

### 2017-08-15 inst/gitbook/02_getting-started.Rmd
book update

# Getting started with dispRity

## What sort of data does dispRity work with?

@@

To illustrate these functions, we will use data from @beckancient2014.
This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).

@@

  * An ordinated matrix (we covered that above)
  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time (i.e. an age for the root), you can simply do `my_tree$root.time <- my_age`.
  * The required number of time subsamples (here `time = 3`)
  * Your favourite disparity metric (here the sum of variances)

Using the @beckancient2014 data described [above](#example-data):

@@

This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:

@@

## Testing for an difference among the time bins
disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
summary(disp_lm)

## Note that ANOVA assumptions were not checked here!
```

@@

Using the @beckancient2014 data described [above](#example-data):

>>>>

# Getting started with `dispRity`

## What sort of data does `dispRity` work with?

@@

To illustrate these functions, we will use data from Beck and Lee (2014).
This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).

@@

  * An ordinated matrix (we covered that above)
  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time (i.e. an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
  * The required number of time subsamples (here `time = 3`)
  * Your favourite disparity metric (here the sum of variances)

Using the Beck and Lee (2014) data described [above](#example-data):

@@

This generates a `dispRity` object (see [here](#guts) for technical details).
When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:

@@

## Testing for an difference among the time bins
disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
summary(disp_lm)
```

@@

Using the Beck and Lee (2014) data described [above](#example-data):

### 2017-08-15 inst/gitbook/03_specific-tutorials.Rmd
book update

@@

For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:

@@

There are many ways of measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994. 
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.

@@

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `variances` | The variance of each dimension | `dispRity` |
2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `mode.val` | The modal value | `dispRity` |

@@

<!-- DATA LOADER -->

@@

@@

## Wilcoxon test applied to time sliced disparity with sequential comparisons, with Bonferroni correction
test.dispRity(disparity_time_slices, test = wilcox.test,
              comparisons = "sequential", correction = "bonferroni")

@@

This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
Let's have a look using our previous example of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.

@@

## Ecological studies of disparity
<!-- NC: I wonder if this might be better just as a "hey we can do ecology stuff too and here is a link to a demo" rather than a new section? The manual is loooooooooong so splitting some stuff out might be good? -->

Disparity analyses are not restricted to palaeobiology analysis.
This package is designed for dealing with ordinated matrices, no matter which data was used in the ordination.
Here is an short example of disparity analysis using ecological data from Deirdre McClean (unpubl.).

The dataset contains an ordinated matrix of 20 dimensions (columns) for 40 elements (rows).
The elements are different experimental plots with discrete variations of nutrient enrichment, depth, and freshwater benthic invertebrates.
<!-- NC: Explain a little more about what the data are here as I don't entirely get it. EG what's a treatement etc? Just a really simple explanataion is perfect -->

```{r}
## Loading the data
data(McClean_data)

## Ordinated matrix
ordinated_matrix <- McClean_data$ordination

## Treatments and depths
treatments <- McClean_data$treatment
depth <- McClean_data$depth
```

### Classic ecological "disparity" analysis

A classical way to represent this ordinated data would be to use two dimensional plots to look at how the different experimental plots differ in their values for nutrient enrichment, depth, and freshwater benthic invertebrates. 
<!-- NC: or similar just to make it clearer -->

```{r, fig.width=6, fig.height=6}
## Setting the colors that will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## Setting the symbols that will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))

## Graphical option
op <- par(bty = "n")
## A classic 2D ordination plot
plot(ordinated_matrix[, 1:2], col = cols, pch = pchs,
     xlab = "PC 1", ylab = "PC 2",
     xlim = range(ordinated_matrix[, 1:2]) + c(0, 100))
par(op)
```

This shows the distribution of the experimental plots along the two first axis of variation of the ordinated distance matrix (i.e. the first two dimensions).
At a first glance, it seems difficult to see a clear effect of the treatments (blue or red) or the depth (rounds or triangles) as the different experimental plots with the same parameters don't seem to cluster together.
However, this may be because this plot ignores the 18 other dimensions of the ordination!
Additionally, these two represented dimensions do not represent a biological reality *per se*; i.e. the values on the first dimension do not represent a continuous trait (e.g. depth), instead they just represent the ordinations of correlations between the data and some factors.

Therefore, we might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a *n*-dimensional object.

### Multidimensional analysis with `dispRity`

#### Splitting the data with respect to the different factors

The first step is to create different subsamples that represent subsamples of the ordinated space (i.e. sub-regions within the *n*-dimensional object).
Each of these subsamples will contain a certain number of elements (i.e. a subset of the 40 experimental field plots) that have some attributes in common.
In our example, we are going to group the elements according to their depth and treatment.

```{r}
## Creating a table that contain the elements and their attributes
factors <- as.data.frame(matrix(data = c(treatments, depth),
          nrow = nrow(ordinated_matrix), ncol = 2, byrow = FALSE,
          dimnames = list(rownames(ordinated_matrix))))
names(factors) <- c("Treat", "Depth")
head(factors)
```

Second, let’s split the data according by depth and treatment to create the subsamples of the ordinated space by using the `custom.subsamples` function:

```{r}
## Splitting the ordinated space into four subsamples
(customised_subsamples <- custom.subsamples(ordinated_matrix, factors))
```

#### Calculating disparity

In this example, we are going to define disparity as the n-dimensional ellipsoid volume (see `?hyper.volume`).

```{r}
## Calculating disparity as the ellipsoid volume of each group
disparity <- dispRity(customised_subsamples, metric = diagonal)
summary(disparity)
```

We can also bootstrap the data to test the robustness of the measured disparity to outliers.
Also, as we can see, each subsample has different numbers of elements.
It might also be interesting to rarefy the data to have only subsamples with the same number of elements.

```{r}
## Bootstrapping the subsamples and using rarefaction
## (i.e. only re-sampling 17 elements each time)
bootstrapped_data <- boot.matrix(customised_subsamples, bootstraps = 100,
                                 rarefaction = 17)
```

We can now run a more robust disparity analysis using the bootstrapped data:

```{r}
## Calculating the bootstrapped disparity
disparity <- dispRity(bootstrapped_data, metric = diagonal)
```

#### Summarising and displaying the results

Now let's summarise this data:

```{r, fig.width=12, fig.height=6, eval=TRUE}
## Summary table of the data
summary(disparity)

## Graphical options
op <- par(mfrow = (c(1, 2)), bty = "n")

## Plotting the bootstrapped disparity
plot(disparity, main = "Bootstrapped", las = 2, xlab = "")

## Plotting the rarefied disparity
plot(disparity, rarefaction = 17, main = "Rarefied", las = 2, xlab = "")

As we can see, there seems to be no strong effect of the number of experimental plots in each subsample (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!

#### Testing our hypotheses

Finally, we can test our hypothesis (whether the water treatment at certain depths alters invertebrate communities and composition in natural habitats) by using the `test.dispRity` function.
<!-- Ah so this is what you are testing! Make this clearer in the intro bit to this section -->

```{r}
## Testing the effect of treatment and depth on the bootstrapped data
summary(test.dispRity(disparity, test = aov, comparisons = "all"))
```

>>>>

Before each section, make sure you loaded the Beck and Lee (2014) data (see [example data](#example-data) for more details).

```{r}
## Loading the data
data(BeckLee_mat50) ; data(BeckLee_mat99)
data(BeckLee_tree) ; data(BeckLee_ages)
```

@@

For the time-slicing method details see Cooper and Guillerme (in prep.). <!-- @@@ Change cite appropriately! -->
<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:

@@

There are many ways of measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space (Donohue _et al._ 2013), the sum and the product of the ranges and variances (Wills _et al._ 1994) or the median position of the elements relative to their centroid (Wills _et al._ 1994). 
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.

@@

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
1 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | Donohue _et al._ (2013) |
1 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
1 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
1 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
1 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `variances` | The variance of each dimension | `dispRity` |
2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `mode.val` | The modal value | `dispRity` |

@@

@@

Here we are using the variables generated in the [section above](#summarising-dispRity-data-plots):

@@

## Wilcoxon test applied to time sliced disparity with sequential comparisons,
## with Bonferroni correction
test.dispRity(disparity_time_slices, test = wilcox.test,
              comparisons = "sequential", correction = "bonferroni")

@@

This is possible in the `dispRity` package by calculating disparity as a dimension-level 2 metric only!
Let's have a look using our [previous example](#summarising-dispRity-data-plots) of bootstrapped time slices but by measuring the distances between each taxon and their centroid as disparity.

@@

### 2017-08-15 inst/gitbook/index.Rmd
book update

This is a package for measuring disparity in `R`.
It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values so they can easily be compared.
This manual is based on the version `1.0`.

@@

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if Claddis is installed BUILD ONLY
# NC: this isn't checking Claddis...
library(dispRity)
## Setting the random seed for repeatability
set.seed(123)
```

@@

[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)

@@

<!-- TG: I'm gonna change that to a newest version with the submission -->

>>>>

This is a package for measuring disparity in `R`.
It allows users to summarise matrices as representations as multidimensional spaces (namely from ordinated matrices from MDS, PCA, PCO, PCoA) into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
This manual is based on the version `0.4`.

@@

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## loading the package and setting up the start seed.
library(dispRity)
## Setting the random seed for repeatability
set.seed(123)
```

@@

<!-- badges out start -->
[![Build Status](https://travis-ci.org/TGuillerme/dispRity.svg?branch=release)](https://travis-ci.org/TGuillerme/dispRity)
[![codecov](https://codecov.io/gh/TGuillerme/dispRity/branch/release/graph/badge.svg)](https://codecov.io/gh/TGuillerme/dispRity)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
<!-- badges out end   -->

@@
